Bias no more: Introduction to latent group-mean centering in hierarchical bayesian models

### 2023-03-17 ms.Rmd
Some notation fixes

\begin{align*}
x_{ij} &\sim \operatorname{Normal}(\alpha_{xj}, \sigma^2_x) \\
y_{ij} &\sim \operatorname{Normal}(
  \alpha_{yj} +
  \beta_{1j}(y_{[i-1]j} - \alpha_{yj}) +
  \beta_{2j}(x_{[i-1]j} - \alpha_{xj}), 
  \sigma^2_y
) \\ \\
\begin{bmatrix} 
  \alpha_x \\ \alpha_y \\ \beta_1 \\ \beta_2
\end{bmatrix} &\sim \operatorname{MVN}(
\begin{bmatrix} 
  \bar{\alpha}_x \\ \bar{\alpha}_y \\ \bar{\beta}_1 \\ \bar{\beta}_2
\end{bmatrix}, \mathbf{\Sigma}) \\ \\
\Sigma &= \mathbf{SRS} \\ \\
\mathbf{S} &= \begin{bmatrix} 
  \tau_0 &0 &0 &0 \\ 
  0 &\tau_1 &0 &0\\ 
  0 &0 &\tau_2 &0 \\ 
  0 &0 &0 &\tau_3
\end{bmatrix} \\ \\
\mathbf{R} &= 
\begin{bmatrix} 
  1 &\rho_{\alpha_x\alpha_y} &\rho_{\alpha_x\beta_1} &\rho_{\alpha_x\beta_2} \\ 
  \rho_{\alpha_x\alpha_y} &1 &\rho_{\alpha_y\beta_1} &\rho_{\alpha_y\beta_2} \\ 
  \rho_{\alpha_x\beta_1} &\rho_{\alpha_y\beta_1} &1 &\rho_{\beta_1\beta_2} \\ 
  \rho_{\alpha_x\beta_2} &\rho_{\alpha_y\beta_2} &\rho_{\beta_1\beta_2} &1 \\
\end{bmatrix}.
\end{align*}

@@

# Save empty brms model
model <- bf(
  y ~ alphay + betayy*(ylag - alphay) + betayx*(xlag - alphax),
  alphax + alphay + betayy + betayx ~ 1 + (1 |i| individual),
  nl = TRUE
) +
  gaussian()
p <- prior(normal(0, 1), class = b, coef = Intercept, nlpar = alphax) +
  prior(normal(0, 1), class = b, coef = Intercept, nlpar = alphay) +
  prior(normal(0, 1), class = b, coef = Intercept, nlpar = betayy) +
  prior(normal(0, 1), class = b, coef = Intercept, nlpar = betayx) +
  prior(lkj(2), class = cor, group = individual) +
  prior(student_t(7, 0, 1), class = sd, group = individual, nlpar = alphax) +
  prior(student_t(7, 0, 1), class = sd, group = individual, nlpar = alphay) +
  prior(student_t(7, 0, 0.5), class = sd, group = individual, nlpar = betayy) +
  prior(student_t(7, 0, 0.5), class = sd, group = individual, nlpar = betayx)
fit_brms_empty <- brm(
  model,
  data = simulate_twolevel_data()$data[[1]],
  prior = p,
  control = list(adapt_delta = 0.95),
  chains = 0,
  file = "models/brm-fit-empty"
)

>>>>

\begin{align*}
x_{ij} &\sim \operatorname{Normal}(\alpha_{xj}, \sigma^2_x) \\
y_{ij} &\sim \operatorname{Normal}(
  \alpha_{yj} +
  \beta_{1j}(y_{[i-1]j} - \alpha_{yj}) +
  \beta_{2j}(x_{[i-1]j} - \alpha_{xj}), 
  \sigma^2_y
)
\end{align*}

The second model equation above specifies that the $\beta$s model effects of the within-person deviated variables of lagged y and x, by virtue of subtracting from each their respective person-specific (latent) mean. I hope. Then we specify the rest of the multilevel model as usual, butchering the notation with bars to indicate average or population-level regression coefficients. We also specify matrices of standard deviations S and correlations R to pop into the covariance matrix (as those will be estimated, rather than covariances, by brms/Stan).

\begin{align*}
\begin{bmatrix} 
  \alpha_x \\ \alpha_y \\ \beta_1 \\ \beta_2
\end{bmatrix} &\sim \operatorname{MVN}(
\begin{bmatrix} 
  \bar{\alpha}_x \\ \bar{\alpha}_y \\ \bar{\beta}_1 \\ \bar{\beta}_2
\end{bmatrix}, \mathbf{\Sigma}) \\ \\
\Sigma &= \mathbf{SRS} \\ \\
\mathbf{S} &= \begin{bmatrix} 
  \tau_{\alpha_x} &0 &0 &0 \\ 
  0 &\tau_{\alpha_y} &0 &0\\ 
  0 &0 &\tau_{\beta_1} &0 \\ 
  0 &0 &0 &\tau_{\beta_2}
\end{bmatrix} \\ \\
\mathbf{R} &= 
\begin{bmatrix} 
  1 &\rho_{\alpha_x\alpha_y} &\rho_{\alpha_x\beta_1} &\rho_{\alpha_x\beta_2} \\ 
  \rho_{\alpha_x\alpha_y} &1 &\rho_{\alpha_y\beta_1} &\rho_{\alpha_y\beta_2} \\ 
  \rho_{\alpha_x\beta_1} &\rho_{\alpha_y\beta_1} &1 &\rho_{\beta_1\beta_2} \\ 
  \rho_{\alpha_x\beta_2} &\rho_{\alpha_y\beta_2} &\rho_{\beta_1\beta_2} &1 \\
\end{bmatrix}.
\end{align*}

We then estimate the resulting datasets with two models. The first is an lmer model that substitutes the observed means $\bar{x}_{.j}$ and $\bar{y}_{.j}$ for the latent intercepts $\alpha_x$ and $\alpha_y$, respectively. The second is our main brms model, specified as follows:

```{r}
#| include: true
#| echo: true
model <- bf(
  y ~ alphay + betayy*(ylag - alphay) + betayx*(xlag - alphax),
  alphax + alphay + betayy + betayx ~ 1 + (1 |i| individual),
  nl = TRUE
) +
  gaussian()
```

Note that this model doesn't estimate an outcome model for $x$, which may or may not have consequences. We then went ahead and ran the simulations with 100 repetitions at each row of the simulation design matrix. This only took a few hours on a M1 laptop. We show results in Figure \@ref(fig:fig-sim-res). First, we see that observed mean centering recovers the specified values of $\alpha_x$ well--those are not in the model output but just means of $\bar{x}_{.j}$. The latent means estimates here are more variable across simulation runs, but also recover the mean well when the number of timepoints is larger than about 10. $\alpha_y$ remains biased upward even with a large number of timepoints per person, but not with the latent means model.

The key coefficients $\beta_1$ and $_2$ show that latent mean centering effectively debiases the autocorrelated parameter, even at low numbers of timepoints. Estimates of $\beta_2$ are noisier across simulation runs but both models seem to recover the values well (observed mean model is consistently greater?)

Other notable findings:
- The correlation between intercepts (between-person correlation) $\rho$ is massively downward biased?
- The between-person standard deviations are all over the place.

@@

# Save empty brms model
p <- prior(normal(0, 1), class = b, coef = Intercept, nlpar = alphax) +
  prior(normal(0, 1), class = b, coef = Intercept, nlpar = alphay) +
  prior(normal(0, 1), class = b, coef = Intercept, nlpar = betayy) +
  prior(normal(0, 1), class = b, coef = Intercept, nlpar = betayx) +
  prior(lkj(2), class = cor, group = individual) +
  prior(student_t(7, 0, 1), class = sd, group = individual, nlpar = alphax) +
  prior(student_t(7, 0, 1), class = sd, group = individual, nlpar = alphay) +
  prior(student_t(7, 0, 0.5), class = sd, group = individual, nlpar = betayy) +
  prior(student_t(7, 0, 0.5), class = sd, group = individual, nlpar = betayx)
fit_brms_empty <- brm(
  model,
  data = simulate_twolevel_data()$data[[1]],
  prior = p,
  control = list(adapt_delta = 0.95),
  chains = 0,
  file = "models/brm-fit-empty"
)
